---
title: "ERHS 535 Final Project"
author: "Rachel Jones"
output: 
  flexdashboard::flex_dashboard:
    theme: bootstrap
---


```{r load_libraries}
library(DT)
library(plotly)
library(readr)
library(tidyverse)

```


```{r reading_in}
# read in data
plants <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-08-18/plants.csv")

```


```{r ISO_codes}
# add ISO country codes for plotly
plants <- plants %>% 
  mutate(code = 
           case_when(country == "Angola" ~ "AGO",
                     country == "Argentina" ~ "ARG",
                     country == "Australia" ~ "AUS",
                     country == "Belgium" ~ "BEL",
                     country == "Bermuda" ~ "BMU",
                     country == "Bhutan" ~ "BTN",
                     country == "Bolivia" ~ "BOL",
                     country == "Brazil" ~ "BRA",
                     country == "Burundi" ~ "BDI",
                     country == "Cabo Verde" ~ "CPV",
                     country == "Cameroon" ~ "CMR",
                     country == "Canada" ~ "CAN",
                     country == "Chile" ~ "CHL",
                     country == "China" ~ "CHN",
                     country == "Colombia" ~ "COL",
                     country == "Congo" ~ "COG",
                     country == "Cook Islands" ~ "COK",
                     country == "Côte d'Ivoire" ~ "CIV",
                     country == "Croatia" ~ "HRV",
                     country == "Cuba" ~ "CUB",
                     country == "Ecuador" ~ "ECU",
                     country == "Ethiopia" ~ "ETH",
                     country == "Fiji" ~ "FJI",
                     country == "France" ~ "FRA",
                     country == "French Polynesia" ~ "PYF",
                     country == "Gabon" ~ "GAB",
                     country == "Greece" ~ "GRC",
                     country == "Guinea" ~ "GIN",
                     country == "Haiti" ~ "HTI",
                     country == "Honduras" ~ "HND",
                     country == "India" ~ "IND",
                     country == "Indonesia" ~ "IDN",
                     country == "Italy" ~ "ITA",
                     country == "Jamaica" ~ "JAM",
                     country == "Kenya" ~ "KEN",
                     country == "Madagascar" ~ "MDG",
                     country == "Malaysia" ~ "MYS",
                     country == "Mauritius" ~ "MUS",
                     country == "Mexico" ~ "MEX",
                     country == "Montserrat" ~ "MSR",
                     country == "Mozambique" ~ "MOZ",
                     country == "Myanmar" ~ "MMR",
                     country == "Namibia" ~ "NAM",
                     country == "New Caledonia" ~ "NCL",
                     country == "New Zealand" ~ "NZL",
                     country == "Nigeria" ~ "NGA",
                     country == "Norfolk Island" ~ "NFK",
                     country == "Panama" ~ "PAN",
                     country == "Papua New Guinea" ~ "PNG",
                     country == "Peru" ~ "PER",
                     country == "Philippines" ~ "PHL",
                     country == "Pitcairn" ~ "PCN",
                     country == "Portugal" ~ "PRT",
                     country == "Rwanda" ~ "RWA",
                     country == 
                       "Saint Helena, Ascension and Tristan da Cunha" ~ "SHN",
                     country == "Sao Tome and Principe" ~ "STP",
                     country == "Seychelles" ~ "SYC",
                     country == "South Africa" ~ "ZAF",
                     country == "Spain" ~ "ESP",
                     country == "Sri Lanka" ~ "LKA",
                     country == "Swaziland" ~ "SWZ",
                     country == "Taiwan" ~ "TWN",
                     country == "Tanzania" ~ "TZA",
                     country == "Trinidad and Tobago" ~ "TTO",
                     country == "Uganda" ~ "UGA",
                     country == "Ukraine" ~ "UKR",
                     country == "United Kingdom" ~ "GBR",
                     country == "United States" ~ "USA",
                     country == "Venezuela" ~ "VEN",
                     country == "Viet Nam" ~ "VNM",
                     country == "Yemen" ~ "YEM",
                     country == "Zimbabwe" ~ "ZWE")) %>% 
  relocate(code, .after = country)

```


```{r data_for_map}
# get country list 
countries <- plants %>% 
  select(country, code) %>% 
  arrange(code) %>% 
  unique()

# add up threats by country
threats <- plants %>% 
  select(country, code, threat_AA:threat_NA) %>% 
  pivot_longer(cols = threat_AA:threat_NA, 
               names_to = "type", values_to = "TF") %>% 
  group_by(code) %>% 
  summarize(n_threats = sum(TF)) %>% 
  ungroup()

# add up actions by country
actions <- plants %>% 
  select(country, code, action_LWP:action_NA) %>% 
  pivot_longer(cols = action_LWP:action_NA,
               names_to = "type", values_to = "TF") %>% 
  group_by(code) %>% 
  summarize(n_actions = sum(TF)) %>% 
  ungroup()

# set up basic descriptive info by country for map
plants_bycountry <- plants %>% 
  group_by(code) %>% 
  summarize(n_plants = n()) %>% 
  ungroup() %>% 
  mutate(country = countries$country) %>% 
  relocate(country, .before = code) %>% 
  full_join(threats, by = "code") %>% 
  full_join(actions, by = "code")

# make column to contain "hover text"
plants_bycountry <- plants_bycountry %>% 
  mutate(hover = with(plants_bycountry, 
                      paste(country, "<br>",
                            "no. plants:", n_plants, "<br>",
                            "no. threats:", n_threats, "<br>",
                            "no. actions:", n_actions)))

```


```{r data_for_bubbleplot}
# add up threats by continent
threats_cont <- plants %>% 
  select(continent, threat_AA:threat_NA) %>% 
  pivot_longer(cols = threat_AA:threat_NA, 
               names_to = "type", values_to = "TF") %>% 
  group_by(continent) %>% 
  summarize(n_threats = sum(TF)) %>% 
  ungroup()

# add up actions by continent
actions_cont <- plants %>% 
  select(continent, action_LWP:action_NA) %>% 
  pivot_longer(cols = action_LWP:action_NA,
               names_to = "type", values_to = "TF") %>% 
  group_by(continent) %>% 
  summarize(n_actions = sum(TF)) %>% 
  ungroup()

# set up basic descriptive info by continent for bubble plot
plants_bycontinent <- plants %>% 
  group_by(continent) %>% 
  summarize(n_plants = n()) %>% 
  ungroup() %>% 
  mutate(n_threats = threats_cont$n_threats) %>% 
  mutate(n_actions = actions_cont$n_actions)

# make column to contain "hover text"
plants_bycontinent <- plants_bycontinent %>% 
  mutate(hover = with(plants_bycontinent,
                      paste(continent, "<br>",
                            "no. plants:", n_plants, "<br>",
                            "no. threats:", n_threats, "<br>",
                            "no. actions:", n_actions)))

```


```{r data_for_tables}
# add up threat types by continent for data table
threats_bycontinent <- plants %>% 
  select(continent, threat_AA:threat_NA) %>% 
  pivot_longer(cols = threat_AA:threat_NA, 
               names_to = "threat_type", values_to = "TF") %>% 
  group_by(continent, threat_type) %>% 
  mutate(threat_total = sum(TF)) %>% 
  slice_head() %>% 
  ungroup() %>% 
  select(-TF) %>% 
  pivot_wider(names_from = continent, values_from = threat_total)

# add up all threats of each type for globe
threats_total <- plants %>% 
  select(continent, threat_AA:threat_NA) %>% 
  pivot_longer(cols = threat_AA:threat_NA, 
               names_to = "threat_type", values_to = "TF") %>% 
  group_by(threat_type) %>% 
  mutate(Total = sum(TF)) %>%
  slice_head() %>% 
  ungroup() %>% 
  select(-TF, -continent)

# pull data together and make threat types more interpretable
threats_bycontinent <- threats_bycontinent %>% 
  full_join(threats_total, by = "threat_type") %>% 
  mutate(threat_type = case_when(threat_type == "threat_AA" ~ 
                                   "agriculture & aquaculture",
                                 threat_type == "threat_BRU" ~ 
                                   "biological resource use",
                                 threat_type == "threat_RCD" ~
                                   "commercial development",
                                 threat_type == "threat_ISGD" ~
                                   "invasive species",
                                 threat_type == "threat_EPM" ~
                                   "energy production & mining",
                                 threat_type == "threat_CC" ~
                                   "climate change",
                                 threat_type == "threat_HID" ~
                                   "human intrusions",
                                 threat_type == "threat_P" ~
                                   "pollution",
                                 threat_type == "threat_TS" ~
                                   "transportation corridor",
                                 threat_type == "threat_NSM" ~
                                   "natural system modifications",
                                 threat_type == "threat_GE" ~
                                   "geological events",
                                 threat_type == "threat_NA" ~
                                   "unknown threat"))

# add up action types by continent for data table
actions_bycontinent <- plants %>% 
  select(continent, action_LWP:action_NA) %>% 
  pivot_longer(cols = action_LWP:action_NA, 
               names_to = "action_type", values_to = "TF") %>% 
  group_by(continent, action_type) %>% 
  mutate(action_total = sum(TF)) %>% 
  slice_head() %>% 
  ungroup() %>% 
  select(-TF) %>% 
  pivot_wider(names_from = continent, values_from = action_total)

# add up all actions of each type for globe
actions_total <- plants %>% 
  select(continent, action_LWP:action_NA) %>% 
  pivot_longer(cols = action_LWP:action_NA, 
               names_to = "action_type", values_to = "TF") %>% 
  group_by(action_type) %>% 
  mutate(Total = sum(TF)) %>%
  slice_head() %>% 
  ungroup() %>% 
  select(-TF, -continent)

# pull data together and make action types more interpretable
actions_bycontinent <- actions_bycontinent %>% 
  full_join(actions_total, by = "action_type") %>% 
  mutate(action_type = case_when(action_type == "action_LWP" ~
                                   "land & water protection",
                                 action_type == "action_SM" ~
                                   "species management",
                                 action_type == "action_LP" ~
                                   "law & policy",
                                 action_type == "action_RM" ~
                                   "research & monitoring",
                                 action_type == "action_EA" ~
                                   "education & awareness",
                                 action_type == "action_NA" ~
                                   "unknown action"))

```


Figures
===============================================================================

Column {data-width = 600}
-------------------------------------------------------------------------------

### Hover over countries for additional info

```{r make_map}
# make interactive map
map <- plot_ly(plants_bycountry, type = "choropleth", 
               locations = plants_bycountry$code, 
               z = plants_bycountry$n_plants,
               text = plants_bycountry$hover, 
               colors = "YlGnBu") %>% 
  colorbar(title = "Total<br>extinct<br>plants",
           orientation = "h",
           y = 0) %>% 
  layout(title = "Global plant extinction<br>(IUCN Red List)")

map

```


Column {data-width = 400}
-------------------------------------------------------------------------------

### Hover over bubbles for additional info

```{r make_bubbleplot}
bubbleplot <- plot_ly(plants_bycontinent, 
                      type = "scatter", mode = "markers",
                      x = ~n_plants, y = ~n_threats, 
                      size = ~n_actions, color = ~continent,
                      text = ~hover, 
                      marker = list(opacity = 0.6, sizemode = "diameter")) %>% 
  layout(title = "Threats and actions by continent",
         xaxis = list(title = "Total number of extinct plants"),
         yaxis = list(title = "Total number of threats"))

bubbleplot

```


### README

**About the data:** This flexdashboard is built from [a Tidy Tuesday dataset contributed by Florent Lavergne](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-08-18/readme.md) that describes extinct plants, ecological threats, and protective actions across the globe. The underlying data comes from the [International Union for Conservation (IUCN) Red List](https://www.iucnredlist.org/). It's important to note that this data comes with many gaps and limitations, but given the vital role of plants in sustaining life on Earth, the data that is available remains very valuable.

**On this tab,** you can explore the global distribution of extinct plants, threats, and actions. The map at left displays total counts of extinct plants at the country level as a scale gradient. The bubble plot above displays the correlation of total threats and total extinct plants at the continent level, with bubble diameter a function of total number of actions. You can navigate the interactive features of these figures as follows:

- *hover:* Put your mouse over a figure element for more information about that element. 
- *toolbar:* When your mouse is over a figure, you can also access a toolbar in the top right corner of the figure that has widgets for adjusting zoom, resetting the view, saving an image, etc. 
- *click and drag:* In the map, this action adjusts which region of the world you are viewing. In the bubble plot, this action zooms in to the rectangular region defined by the diagonal line you "draw".

**On the second tab** (accessible via the top-of-page toolbar), you can explore the types of threats facing plants, and the types of actions taken to conserve them, on a per-continent basis. You can navigate the interactive features of these figures as follows:

- *sort*: You can use the "up" and "down" arrows beside each column header to sort the table in ascending or descending order of the column variable.
- *search:* You can search each table for a particular term.


Tables {data-orientation=rows}
===============================================================================

Row
-------------------------------------------------------------------------------

### Threats

```{r make_threat_table}
# make table of threat types by continent
threat_dt <- datatable(threats_bycontinent,
                       colnames = c("Threat type",
                                    "Africa",
                                    "Asia",
                                    "Europe",
                                    "North America",
                                    "Oceania",
                                    "South America",
                                    "Total"),
                       rownames = FALSE,
                       extensions = "Scroller",
                       width = "100%",
                       options = list(deferRender = TRUE, 
                                      scroller = TRUE,
                                      scrollY = 300))

threat_dt

```


Row
-------------------------------------------------------------------------------

### Actions

```{r make_action_table}
# make table of action types by continent
action_dt <- datatable(actions_bycontinent,
                       colnames = c("Action type",
                                    "Africa",
                                    "Asia",
                                    "Europe",
                                    "North America",
                                    "Oceania",
                                    "South America",
                                    "Total"),
                       rownames = FALSE,
                       extensions = "Scroller",
                       width = "100%",
                       options = list(deferRender = TRUE, 
                                      scroller = TRUE,
                                      scrollY = 300))

action_dt

```

